\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-01-30 paper/main.tex

\todo[inline]{First part is not about evaluation but optimization
  objectives, should go in the Model section:}
As the number of drug-targets are sparsely given w.r.t.\todo{no
  acronyms} to the number of both drugs and proteins considered, the
resulting training, validation and testing datasets are highly
imbalanced. As there are only $~22.000$\todo{exact number} links in
the considered subset the ratio
\begin{equation*}
	\frac{\#dti\_links}{\#drugs \cdot \#proteins} \approx 360,
\end{equation*}
consequently needing compensation in the computed loss function and
appropriate metrics for the evaluation.

@@

\subsection{Choosing a cross-validation splitting scheme}

% As in machine learning inference is derived from the underlying data,
% models and data are naturally and intrinsically linked. Thus, the more
% we understand about the pitfalls and biases within the data, the more
% we can try to bypass these difficulties. We will hereby abbreviate,
% e.g., a cross validation splitting scheme as \glqq split\grqq{}
% determining the train, validation and test subset of a given dataset.

Within DTI prediction, one can observe basic biases resulting from the
underlying datasets \citep{Pahikkala2014}. First, novel drugs are
often designed by altering non-functional components of a drug,
leading to two and more very similar drugs designed to target the same
proteins \cite{Overington2006}. This can lead to a bias when it leads
to \glqq hidden duplicates\grqq{} that can distribute among the
train--test split, resulting in a better (measured) predictive
performance than would be expected when the model is applied to
genuinely unknown interactions. Second, some proteins (which we will
denote as \textit{hub proteins}) have significantly more known
interactions with drugs than others. In the STITCH database, $5\%$
of the proteins have $40\%$ of the interactions, and similar
distributions are present in the Yamanishi and Drugbank datasets
\cite{Drugbank2007, Drugbank2017} datasets; preferentially predicting
these proteins can often increase predictive performance while again
not reflecting the actual performance when applied to a new protein
(e.g., a protein for which no interactions are known). This bias in
the number of drugs targeting certain proteins may be the result of
study bias where more \glqq valuable\grqq{} proteins have more drugs
designed to target them due to their involvement in more common
diseases (or diseases for which drugs can be more profitably
marketed).

The aim of \name is to predict candidate drugs that target a given
protein; the challenge is to develop a training and evaluation scheme
that does not simply overfit to the inherent biases in training and
testing data.
In general, when performing cross-validation for DTI prediction, the
options are to split over drug--target pairs, split over drugs, and
split over proteins.

where the first and third option concern splitting drugs and proteins, respectively, into train, validation and test sets, and arranging the corresponding drug-target interactions. Hereby, different training and prediction schemes lead to divergent expressiveness of the resulting model. \\
The most common scheme for DTI prediction is the split over drug-target pairs \citep{Survey2018}, where likely all drugs and targets of the validation and testing phase have already occurred in the training phase, as part of other drug-target pairs. Thus, the first and second splitting scheme are exposed to the first dataset bias and are hence more likely vulnerable to transductive inference by just predicting recently seen structures, rather than implementing inductive inference and generalizing over the drug representations. Second, these two strategies are more susceptible to the second and third bias, as only in these cases the model may overfit on the number of existing interactions, while in the third scheme the number of interactions of the test proteins is not known. \\

Considering the publications of the last

In general, drug-target interaction prediction is the task of accurately predicting, whether for a given drug and a given protein there is a biological interaction within the target organism. Hereby, different training and prediction schemes lead to divergent expressiveness of the resulting model. However, when building the train-test split over compound-protein pairs for building the actual model, there are the following three options:

\todo[inline]{Relate these to the issues listed above:}
\begin{enumerate}
	\item Build split over drugs
	\item Build split over drug-target pairs
	\item Build split over proteins
\end{enumerate} 
In general, recent works do perform their split over the drugs or drug-target pairs (\cite{Survey2018}, CITATION). As there are hopefully many more drugs to discover, the drug split scheme both emphasizes the drug repurposing idea, by applying unseen compounds to existing targets, but also benefits from more complicated drug representations, leading to tremendous\todo{???} results. This performance gain is based on minor variations among large groups of pharmaceuticals, that are easy to acquire\todo{evidence}. The second scheme has knowledge on all drugs and all proteins, and is thus prone to overfitting and the same development bias. Eventually, as there only limited drug-targets \citep{Overington2006}, predicting per protein is rather counter-intuitive. As it is hard to generalize over proteins representations, we aim at reaching similar performances for both drug and protein splitting schemes. \\

In general, recent works do perform their split over the drugs or drug-target pairs (\cite{Survey2018}, CITATION). The first is more relevant for novel drugs, as it is much more likely to test a new compound than a innovative protein. However, it lies in the very nature of the used datasets, making the prediction for new drugs much easier. Thus, drugs are often built by minor variations of existing drugs, thus leading to no deviations in the functional group of that very compound (CITATION/EXAMPLE). When distributed over both train and test split, the models do not perform inductive inference and generalize, but rather implement transductive inference by just predicting the recently seen structures. Hence, when entirely new molecules are seen, the models perform much worse

The same applies to splits of drug-target pairs, as all drugs were already seen, and novelty cannot be coped with.

As mentioned in the introduction it is quite difficult to learn suitable features from proteins. In general, attempts search for motifs in the protein sequences under usage of convolutional neural networks and filters, which is more suitable for tasks like protein function prediction, than for for drug-target interaction prediction, and lack a more in-depth hypothesis on the protein side, while investing in refined drug features. \\
Thus, building splitting over proteins is the most challenging of the three options. \\

>>>>

\todo[inline]{First part is not about evaluation but optimization
  objectives, should go in the Model section:}
As the number of drug-targets are sparsely given w.r.t.\todo{no
  acronyms} to the number of both drugs and proteins considered, the
resulting training, validation and testing datasets are highly
imbalanced. As there are only $~22.000$\todo{exact number} links in
the considered subset the ratio
\begin{equation*}
	\frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
\end{equation*}
consequently needing compensation in the computed loss function and
appropriate metrics for the evaluation.

@@

\subsection{Choosing a cross-validation splitting scheme}
As in machine learning inference is derived from the underlying data, models and data are naturally and intrinsically linked. Thus, the more we understand about the pitfalls and biases within the data, the more we can try to bypass these difficulties. We will hereby abbreviate, e.g., a cross validation splitting scheme as \glqq split\grqq{} determining the train, validation and test subset of a given dataset. \\
Especially within DTI prediction, one can observe elemental and basic skews \citep{Pahikkala2014}. (1.) Frequently, novel drugs are designed by altering non-functional components of a drug, leading to two and more very similar drugs, designed to target the same proteins (\textbf{CITATION?}). This bias can lead to \glqq hidden duplicates\grqq{} that can distribute among the train-test split, skewing the ultimate predictive performance. (2.) Additionally, within the realm of DTI prediction, some proteins, which we will denote as \textit{hub proteins}, have significantly more known interactions with drugs than others. In the considered subset of STITCH database, $5\%$ of the proteins had $40\%$ of the interactions, with similar biases for Yamanishi and Drugbank (\citet{Drugbank2007} and \citet{Drugbank2017}) datasets, making them easy and reliable predictors. (3.) The second distortion could be based on a publication bias for more \glqq valuable\grqq{} proteins due to their disease involvement, such as e.g., cancer, leading to more research and eventually more interacting drugs regarding those. Thus if we aim to find a drug for a given protein, the challenge is to develop a training and evaluation scheme, that does not simply overfit to those biases.\\

In general, when concerning cross-validation splitting schemes for DTI prediction, there are the following three options:
\begin{enumerate}
	\item Build split over drugs
	\item Build split over drug-target pairs
	\item Build split over proteins
\end{enumerate}
where the first and third option concern splitting drugs and proteins, respectively, into train, validation and test sets, and arranging the corresponding drug-target interactions. Hereby, different training and prediction schemes lead to divergent expressiveness of the resulting model. \\
The most common scheme for DTI prediction is the split over drug-target pairs \citep{Survey2018}, where likely all drugs and targets of the validation and testing phase have already occurred in the training phase, as part of other drug-target pairs. The second most prevalent arrangement is the split over drugs, while only close to none is aiming on a protein split.  However, the first and second splitting scheme are exposed to the first dataset bias and are hence more likely vulnerable to transductive inference by just predicting recently seen structures, rather than implementing inductive inference and generalizing over the drug representations. Second, these two strategies are more susceptible to the second and third bias, as only in these cases the model may overfit on the number of existing interactions for a single protein, while in the third scheme the number of interactions of the test proteins is not known.

Assuming a hypothetical, perfectly generalizing model built upon an unbiased dataset, this very model would yield similar performances for all three splitting schemes, not overfitting on the known structures. On the other hand, for a hypothetical entirely overfitting model, trained on a highly biased dataset, this model would show substantial deviations from the original performance over another split. \\
We emphasize, that all real-world models are prone to some sort of overfitting, and unknown, deviant entities in both validation and testing set will likely lead to some sort of performance gap for relevant metrics. However, a large disparity may hint the biases stated above.

Hence, we will perform a cross-validation over proteins for the training and prediction phase of our model, despite predicting per protein is rather counter-intuitive as there only limited drug-targets \citep{Overington2006}, and thus novel drugs are more likely to arise than novel targets. Yet, we aim to find all interacting drugs for existing targets motivating our split choice even further.

\subsection{MicroAUC}
Continuing this line, we propose the well known metric \textit{MicroAUC} for a better evaluation for this purpose. For given lists $D, P$ of drugs and proteins, respectively, and a set of known interactions $Int := \{(d_i, p_i) \}$, \textit{MicroAUC} is calculated as the average per entity (macro) \textit{AUROC} score. With respect to proteins, this can be formalized for given labels and prediction $l,y:D\times P \rightarrow \{0,1\}$  as 
\begin{equation*}
	MicroAUC_p := \underset{p\in P}{mean}\left(\left\{ \text{AUROC}(\{ (l(d_i, p), y(d_i,p))| d_i\in D\}) \right\}\right)
\end{equation*}

Note that drugs and proteins can be interchanged in this formulation, which we will denote with $MicroAUC_p$ and $MicroAUC_d$, respectively. As we want to evaluate the models performance to find all suitable drugs for each protein individually, \textit{$MicroAUC_p$} seems to be more reasonable in comparison to both $MicroAUC_d$ and $AUROC$ with respect to the previously proposed protein split cross-validation. Furthermore,
